Discussion Draft

### 2020-04-06 doc/Discussion.rmd
futial effort from last night

Our results of greater between cell variation in females, these results are supported by human and mouse [@lynn2002, @gruhn2013].

While the **modifier models** parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

In the **spindle based** selection model, the less efficient SAC in oocytes could lead to relaxed selection on the chromosome structure. If the SAC/checkpoint which normally acts in a stabilizing selection on the chromosome structure has a reduction in efficiency (relaxed selection), will result in increased variance in (these associated traits), not just across strains (populations), but across cells within individuals (quant-gen ref?). Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers (aMTOCs) (Schuh Ellenburg 2007) which result in

<stabilization -- bi-oriented, robust SAC response, SAC acts to correctly time metaphase to anaphase transition -- only after correct alignment and biorientation of bivalents in the metaphase cell> -- spindle dynamics much different due to asymetrical diviation Mogessie et al 2018, Mogessie Schuh --actin and MT are apart of oocyte spindle.

- which might have a consequence of reducing the (reaction/signal of the SAC (anaphase promoting complexes (cite) )).  (this is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite))

<Bennabi et al 2016 --spindle assembly inoocytes, sun and kim 2011 -- SAC and regulators, Kolano et al 2012 --error prone oocytes, Manil-Segalen -- forces through MT breaking chromsomes>
Mogessie et al 2018 -- meiotic spindle (big)review>
(what generates MT nucleation is unclear (aMTOC or chromsoomes))

@@

Our results on the broad scale pattern of crossover placement are supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait.

The **two locus modifier** and **SACE** model make predictions for sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks respectively.

The two mechanical models also both predict sexual dimorphism in the recombination landscape for distinct reasons. The COM model (chromosome pairing based model), predicts the sex differences in the positioning of crossovers is due to a combination of the length of the axis and intensity of the rapid prophase movements (RPMs) [@hulten2011_COM, @rubin2020mixing].

Under the **spindle based model** we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (cite). The number and placement of crossovers alter the distribution of sister cohesion and the resulting 'chromosome structure' when bivalents are aligned on the metaphase plate (cite). The male crossover telomere bias could be adaptive in order to synchronize (or minimize) the timing of homologs separating at metaphase I.	Differing lengths of the timing of cell cycle between oogenesis and spermatogenesis may impose the the different selective pressures. Faster spermatogenesis may select for synchronization of the separation homologs or faster separatation at the metaphase to anaphase transition. While in oogenesis, the slower cell cycle and prolonged arrest may require chromosome structures with greater stability on the MI spindle.

@@

**we find support that interference strength has evolved in the two groups of male strains. There is a small number of positive correlations between genome wide recombination rate and interference strength in the literature.**

@@

Logically a negative correlation is expected; increasing the number of crossovers across chromosomes would most logically be done by  more densely spacing crossovers along chromosomes and decreasing interference strength. This pattern has empirical support from the most species [@ottoPaysuer2019] and fits well with  the fundamental relationship between the SC area or axis length, the physical upper limit for the number of crossovers. **although it is not particulary strong in our results** We argue that this is due to difference in scale (genome wide vs single chromosome)

Our results support an opposite pattern predicted from the 'logical model', a positive correlation between interference strength and genome wide recombination rates; The within sex comparison of two breeds of cattle with different genome wide recombination rates  [@ma2015_cattle], between lab and wild mice of Peromyscus leucopus, [@peterson2019].

Theoretical models haven't really considered the evolution of interference strength, neither the haploid selection or two locus modifier model can not be applied to evolution of interference strength. While the SACE modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a landscape with stronger interference strength. The COM model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, it lacks evolutionary based predictions. We propose the spindle based selection model would support the evolution of interference strength in the positive direction via modulation of the amount of sister cohesion connecting homologs (figure).

@@

Biological exceptions to interference:  experimentally increasing crossover number through mutants or fusion chromosomes  (Celegans, (plants, fungi with negative interference.

>>>>

Our results of greater between cell variation in females, these results are supported by human and mouse [@lynn2002, @gruhn2013]. While the **modifier models** parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

In the **spindle based** selection model, the less efficient SAC in oocytes leads to relaxed selection on the chromosome structure due to meiotic progression even with with unstable bivalents. This will result in increased variance in chromosome structure, and crossover number, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers (aMTOCs) (Schuh Ellenburg 2007). This is thought to reduce the triggering signal for the SAC.  This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite)).

<stabilization -- bi-oriented, robust SAC response, SAC acts to correctly time metaphase to anaphase transition -- only after correct alignment and biorientation of bivalents in the metaphase cell, -- spindle dynamics much different due to asymetrical diviation Mogessie et al 2018, Mogessie Schuh --actin and MT are apart of oocyte spindle>

<Bennabi et al 2016 --spindle assembly inoocytes, sun and kim 2011 -- SAC and regulators, Kolano et al 2012 --error prone oocytes, Manil-Segalen -- forces through MT breaking chromsomes, Mogessie et al 2018 -- meiotic spindle (big)review>

@@

Our results on the broad scale pattern of crossover placement are supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The **two locus modifier** and **SACE** model predicts sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The two mechanical models both predict sexual dimorphism in the recombination landscape for distinct reasons. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differening strengths of chromosome movements during the pairing process [@hulten2011_COM, @rubin2020mixing].

Under the **spindle based model** we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (cite). The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate (cite). Different timing of cell cycle between oogenesis and spermatogenesis may impose the the different selective pressures on sister cohesion. Faster spermatogenesis may select for synchronization of the separation homologs. While in oogenesis, the slower cell cycle and mulitple arrest stages may require chromosome structures with greater stability on the MI spindle.

<The male crossover telomere bias could be adaptive in order to synchronize (or minimize) the timing of homologs separating at metaphase I.>

@@

@@

Logically a negative correlation is expected; increasing the number of crossovers across chromosomes would most logically be done by  more densely spacing crossovers along chromosomes and decreasing interference strength.

This pattern has empirical support from the most species [@ottoPaysuer2019] and fits well with  the fundamental relationship between the SC area or axis length, the physical upper limit for the number of crossovers. (**although it is not particulary strong in our results**).

**interference strength evolve**
**other examples of interferene strength evolution, tricky due to chromosome specific effects and complications from large evolutionary distance (change in underlying karyotype**

Our results find that interference strength has evolved in males from 3 strains from our panel of house mice.

Examples of evolution of interference strength are rare and complicated by changes in the underlying karyotype structure (Segura et al 2014).

The general patern of the relationship
 -- most findings / the general pattern of interference evolution is that it has a negative correlation to genome wide recombination rate. (Review in otto payseur), X examples -- correlation of X across x data.
--This fit logic of XX -- and also can be applied to the known mechanism of DSB formation / and the relationship with chromosome axis.

While examples of positive correlation of interference strength and genome-wide recombination rate are less common; the within sex comparison of two breeds of cattle with different genome wide recombination rates  [@ma2015_cattle], between lab and wild mice of Peromyscus leucopus, [@peterson2019].

we propose that --- this is due to difference in scale (resolution)
and how these two metrics are measured at the genome wide vs single chromosome (bivalent) level.

<explain this difference>

<**the models**>

The **COM** model predicts that interference and the recombination landscape arises from known oscillatory movements during prophase, it lacks evolutionary based predictions.

While the **SACE** modifier model does not explicitly model evolution of interference strength we note that a logical outcome of the main prediction of maintaining larger chromosome blocks in males, would be a landscape with stronger interference strength.

 We propose the **spindle based selection model** would support the evolution of interference strength in the positive direction via modulation of the amount of sister cohesion connecting homologs (figure).

<**other models**>

@@

<Biological exceptions to interference:  experimentally increasing crossover number through mutants or fusion chromosomes  (Celegans, (plants, fungi with negative interference.>
